---
title: "Assessment"
author: "Eric Odongo"
date: "2025-02-22"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Part 1: Data Processing

**Tasks**

1. Load the dataset and cast it into a data.table.

```{r}

library(nycflights13)
library(data.table)
library(RSQLite)


#load the dataframe and convert to a data.table

data('flights')

flightsDT <- as.data.table(flights)

class(flightsDT)

```


2. Perform the following transformations:

– Compute the average departure delay for each airline.

```{r}

#get the airlines data set and merge on flights data set to extract the airline names

data('airlines')

airlinesDT <- as.data.table(airlines)

flightsDT[airlinesDT, on = 'carrier', airline_name := i.name]


#summarize by airline

flightsDT[, mean(dep_delay, na.rm = T), by = airline_name] |> 
  knitr::kable(col.names = c('Airline', 'Average Departure Delay'),
               digits = 3)


```


– Find the top 5 destinations with the most flights.

```{r}

#get airport names

data('airports')

airportsDT <- data.table(airports)

setnames(airportsDT, 'faa', 'dest')

flightsDT[airportsDT[, c('dest','name')], on = 'dest', airport_name := i.name]


top_dest <- flightsDT[, by = airport_name, .N][order(-N)][1:5]

top_dest |> 
  knitr::kable()


```

Top five airport destinations with the most flights are Chicago Ohare Intl, Hartsfield Jackson Atlanta Intl, Los Angeles Intl, General Edward Lawrence Logan Intl, and Orlando Intl.

\newpage

– Create two new columns:

• one which is a unique id for each row/flight.

```{r}

flightsDT[, id := .I]

```


• and another indicating whether a flight was delayed for more than 15
minutes.

```{r}

flightsDT[, indicator := ifelse(dep_delay > 15, 'delayed', 'not delayed')]

table(flightsDT$indicator)

```



3. Save the processed data as a CSV or an SQLite database file.

```{r}

write.csv(flightsDT, 'Data/flights_final.csv')

#save to a database file

# Set up database

drv <- dbDriver("SQLite")
tfile <- "Data/nycflights.db"

con <- dbConnect(drv, dbname = "Data/nycflights.db")

dbWriteTable(con, "flightsDT", as.data.frame(flightsDT))
dbDisconnect(con)

```

\newpage

## Part 2: JSON Data API

